Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-25 ms.tex
organizing a bit

\subsection{Version control}

@@

% With suggested edits, some of this moves up.
Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}. \texttt{git} is attractive for data management because it tracks all changes in monitored files. This history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN here?
That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

\subsection{`Semantic' versioning}

% WKC: this is back to our main point

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

\subsection{Digital object identifiers}

@@

% The introduction talks at length about medium sized data and little about versioning but this section talks only about versioning

@@

%work on this paragraph more
The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.  Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but the current structure of scientific research is very distributed, with research groups across the world working on the same question.  Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system. The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.

The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s.  Moreover, it is relatively easy to update and add new versions.  Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.  But, nonetheless, workable systems are already possible.

% potentially move to discussion? 
\subsection{Cloud platforms for storing and distributing version controlled data}

There are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.

% Though this section I am losing track of the big picture here;
% there's lots of discussion of the positive benefits of "Something"
% but without knowing what it is, it's unlikely that these will stick.
% I wonder if we should move the workflow description higher.

% WKC: I agree. Not sure what to do with these two sections which are a bit of a tangent relative to our main point

% potentially move to discussion?
\subsection{Potential for collaboration: allowing for multiple data contributors}

One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. This is enabled via the distributed and decentralised nature of \texttt{git} repositories. Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
% We have constant drama at work because the different spreadsheet
% writers we use across the platforms quote files inconsistently so
% every save is a complete rewrite of the file.  It's totally
% impossible to do any collaborative editing....
 or, when needed, flag where there are conflicts that need to be resolved. This makes collaboration easy. For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project. This process is called submitting a ``Pull request''. One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
 % This section conflates git and github more than is useful.  It
 % might be worth setting up github as the explicit platform here or
 % dialing back the language a bit.  GitLab and bitbucket have their
 % own "merge request" feature (I think).  It's also github that makes
 % repositories _discoverable_ which is a requirement for getting
 % outside collaborators in.  Oddly the decentralised nature of git is
 % not actually that useful in the github model because github is so
 % strongly centralised.  Will and I did some true decentralised git
 % collaboration on a plane to nescent once with a bare git repo on a
 % thumb drive.

>>>>

\subsection{Version control and `semantic' versioning}

@@

% With suggested edits, some of this moves up.

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}. \texttt{git} is attractive for data management because it tracks all changes in monitored files. This history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it \emph{might} change the results of some users' analyses. A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.  
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

@@

% The introduction talks at length about medium sized data and little about versioning but this section talks only about versioning
% WKC: now intro is about versioning and data size issue discussion moved here, which I think is an improvement

@@

%work on this paragraph more
The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.  Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but, while there are cases where data curation may be done by a large institution, the current structure of most scientific research is very distributed, with a few research groups across the world working on similar questions.  Moreover, curation of data on particular questions requires, in many cases, a great deal of expertise in the field, so the experts in the field need to be involved in data curation. However, individual research groups have different capabilities compared to institutional data curators: at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system. There are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.

The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Figure \ref{fig:technology_stack}).  Moreover, it is relatively easy to update and add new versions.  Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.  But, nonetheless, workable systems are already possible.

One of the greatest benefits of using a cloud--based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. This makes collaboration easy.

\subsection{Future directions}

Technology for data management and distribution is moving forward rapidly.  The workflow we describe above works well with current tools and should work even more quickly with its future replacement.  Here we highlight a few rapidly developing future tools that researchers should be aware of.

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN here?
That said...
% Add CKAN and Git-LFS, and say something useful about the future

% We have constant drama at work because the different spreadsheet
% writers we use across the platforms quote files inconsistently so
% every save is a complete rewrite of the file.  It's totally
% impossible to do any collaborative editing....
% For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project. This process is called submitting a ``Pull request''. One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
 % This section conflates git and github more than is useful.  It
 % might be worth setting up github as the explicit platform here or
 % dialing back the language a bit.  GitLab and bitbucket have their
 % own "merge request" feature (I think).  It's also github that makes
 % repositories _discoverable_ which is a requirement for getting
 % outside collaborators in.  Oddly the decentralised nature of git is
 % not actually that useful in the github model because github is so
 % strongly centralised.  Will and I did some true decentralised git
 % collaboration on a plane to nescent once with a bare git repo on a
 % thumb drive.

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
